---
title: "lab1"
author: "Brian Franco"
date: "2024-08-29"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary: Chapter 1 Summary
Chapter 1:
The organization of data plays a crucial way with how we analyze and interpret information. When it comes to an elastic method to organize data we can often rely on whats called a data matrix. A data matrix proves to be reliable especially for spreadsheets where a row fits to an observational unit, and each columns matches up to a given variable. Variables can be broken into to types, numerical and categorical. Numerical can be split into two sub groups, discrete and continues. Discrete variables only assume values that are countable whereas continues variables are uncountable. Categorical also breaks down into two distinctive sub groups, nominal and ordinal. For Nominal variables the order for them does not matter but for ordinal the order does. When it comes to data collection there are two main types, observational studies and experiments. Observational studies collect data such that they don't interfere with the data analyzed, and experiments researches the possibility of a casual relationship where there will exists both response and explanatory variables. Data that is observed and nothing is applied is called observational data. There exists 4 sampling methods for random sampling techniques: simple random, stratified, cluster, and multistage sampling. Simple random sampling is when every single case has a likely chance of being selected. Stratified sampling is when a similar set are grouped together, typically each group is sampled using random sampling within each stratum. A cluster sample we split the population into a few groups, we take a sample from within the groups often a fixed number and maintain a sample from each of the clusters. A multi stage sample is similar to that of a cluster sample, but instead we take a random sample from within each cluster.


Chapter 2:
Using various methods such as scatter plots, dot plots, histograms, box plots, we can successfully examine numerical data to visualize a given result. 




\newpage 
# Lab One Exercise Chapter 1
```{r, include=FALSE}
source("http://www.openintro.org/stat/data/arbuthnot.R")
```
## Exercise 1

### What command would you use to extract just the counts of girls baptized? Try it!

To retrieve only the counts of girls baptized:
```{r echo=TRUE,results='hide'}
arbuthnot$girls
```

## Exercise 2

### Is there an apparent trend in the number of girls baptized over the years? How would you describe it?

There is a noticeable upward trend in the number of girls baptized starting from 1660 onwards. This indicates a positive correlation between the years and the number of girls baptized after that period.


## Exercise 3

### Now, make a plot of the proportion of boys over time. What do you see? 

From the plot, it appears that the number of baptized boys has also risen after 1660, although the boys’ chart exhibits more fluctuations compared to the girls’ chart.
```{r echo=FALSE, results='hide'}
plot(arbuthnot$year,arbuthnot$boys, type = "l")
```
\newpage
# Lab One OnYourOwn Chapter 1
```{r, include=FALSE}
source("http://www.openintro.org/stat/data/present.R")
```

### What years are included in this data set? What are the dimensions of the data frame and what are the variable or column names?

This dataset spans from 1940 to 2002. The data frame consists of 63 rows and 3 columns. The columns are named “year”, “boys”, and “girls”.

### How do these counts compare to Arbuthnot’s? Are they on a similar scale?


The numbers in this dataset are significantly higher but maintain a similar scale to those in Arbuthnot’s data.

### Make a plot that displays the boy-to-girl ratio for every year in the data set. What do you see? Does Arbuthnot’s observation about boys being born in greater proportion than girls hold up in the U.S.? Include the plot in your response.


The plot suggests that Arbuthnot’s observation holds true to some extent in the U.S., as the points representing boys and girls each year generally lie to the right of the vertical axis, indicating a higher number of boys each year.
```{r echo=FALSE}
plot(x = present$boys, y = present$girls)
```


### In what year did we see the most total number of births in the U.S.?

The highest number of births occurred in 1961, with a total of 4,268,326 births.

\newpage

# Lab One Exercise Chapter 2
```{r,include=FALSE}
source("http://www.openintro.org/stat/data/cdc.R")
```
### How many cases are there in this data set? How many variables? For each variable, identify its data type (e.g. categorical, discrete).

The dataset contains 20,000 cases and includes 9 variables. Variables such as exerany, hlthplan, and smoke100 are discrete numerical variables, whereas height, weight, wtdesire, and age are continuous numerical variables. Additionally, genhlth and gender are categorical variables.


### Create a numerical summary for height and age, and compute the interquartile range for each. Compute the relative frequency distribution for gender and exerany. How many males are in the sample? What proportion of the sample reports being in excellent health?

A summary of the height variable :
```{r echo=FALSE}
summary(cdc$height)
```
The interquartile range for height is 26.

A summary of the age variable :
```{r echo=FALSE}
summary(cdc$age)
```
The interquartile range for age is 6.

The relative frequency distribution for Gender :
```{r echo=FALSE}
table(cdc$gender)/20000
```
The relative frequency distribution for Exerany :
```{r echo=FALSE}
table(cdc$exerany)/20000
```

There are 9,569 males in the sample, and 23% of the participants report being in excellent health.

### What does the mosaic plot reveal about smoking habits and gender?

The mosaic plot indicates that a higher proportion of males smoke compared to females within the sample.

### Create a new object called under23_and_smoke that contains all observations of respondents under the age of 23 that have smoked 100 cigarettes in their lifetime. Write the command you used to create the new object as the answer to this exercise.
The command to create the new object:
```{r,results='hide'}
under23_and_smoke <- subset(cdc, age < 23 & smoke100 == 1)
```

### What does this box plot show? Pick another categorical variable from the data set and see how it relates to BMI. List the variable you chose, why you might think it would have a relationship to BMI, and indicate what the figure seems to suggest.

The box plot comparing BMI and general health (genhlth) illustrates the distribution of BMI across different levels of general health, highlighting their association. I selected the smoke100 variable because I hypothesized that smoking might impact BMI due to its effects on metabolism and overall health. The box plot suggests that there is no significant difference in BMI between smokers and non-smokers.

\newpage

# Lab One OnYourOwn Chapter 2

### Make a scatterplot of weight versus desired weight. Describe the relationship between these two variables.
```{r echo=FALSE}
plot(cdc$weight,cdc$wtdesire)
```

The scatterplot shows that individuals weighing 200 pounds or less generally desire a weight close to their actual weight. In contrast, those weighing over 200 pounds tend to aim for a lower desired weight compared to their current weight.

### Let’s consider a new variable: the difference between desired weight (wtdesire) and current weight (weight). Create this new variable by subtracting the two columns in the data frame and assigning them to a new object called wdiff.
```{r}
wdiff <- (cdc$wtdesire - cdc$weight)
```

### What type of data is wdiff? If an observation wdiff is 0, what does this mean about the person’s weight and desired weight. What if wdiff is positive or negative?

wdiff is an integer variable, representing numerical discrete data. If wdiff is 0, it means the individual’s desired weight is equal to their current weight. A positive wdiff indicates that the desired weight is higher than the current weight, while a negative wdiff means the desired weight is lower than the current weight.

### Describe the distribution of wdiff in terms of its center, shape, and spread, including any plots you use. What does this tell us about how people feel about their current weight?

```{r echo=FALSE}
hist(wdiff)
```

The distribution of wdiff is centered around -14, is unimodal and skewed, with a somwhat narrow spread. This suggests that most people desire to weigh less than they currently do, but the desired reduction is generally moderate rather than extreme.

### Using numerical summaries and a side-by-side box plot, determine if men tend to view their weight differently than women.

Men tend to have different perceptions of their weight compared to women. The minimum desired weight for males is 77, while for females it is 68. The maximum desired weight for males reaches 680, compared to 350 for females. Additionally, males have higher median and mean desired weights than females. When comparing the differences between actual and desired weights side by side, the distributions are largely similar, though males show a tendency to desire a slightly higher weight than females. This indicates that males generally aim for a higher body weight than females.

### Now it’s time to get creative. Find the mean and standard deviation of weight and determine what proportion of the weights are within one standard deviation of the mean.

The average weight(mean) is 169.7 pounds with a standard deviation of 40.1 pounds. Approximately 71% of the weights fall within one standard deviation of the mean.